---
title: "County Level Turnouts and Graphs"
author: "Theodore Dounias"
date: "September 1, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(knitr)
library(plotly)
library(maps)
opts_knit$set(root.dir = 'C:/Users/tdounias/Desktop/Reed College/2017 Fall/MATH 243/Repos/Reed_Senior_Thesis/Data_and_results/')
```
   
#Reading in the Data
```{r, warning=FALSE, error = FALSE}
#Read in voter history data
vhist <- read_csv("data/2017_CO/VHist_2017/CO_2017_VHist_full.csv", 
    col_types = cols(PARTY = col_skip(), 
        VOTER_ID = col_skip(), X1 = col_skip()))

#Read in vrf data
vrf <- read_csv("data/2017_CO/VRF_2017/CO_2017_VRF_full.csv", 
    col_types = cols_only(VOTER_ID = col_guess(), COUNTY = col_guess(), REGISTRATION_DATE = col_guess(), 
                          EFFECTIVE_DATE = col_guess(), VOTER_STATUS = col_guess(), PARTY = col_guess()))
```
   
#Sidenote: Problems in VHist I did not initially find
   
I include this here as a reminder and description of the problem.
```{r}
summary(as.factor(vhist$VOTING_METHOD))

vhist_na <- vhist %>%
  filter(is.na(VOTING_METHOD))

summary(as.factor(vhist_na$COUNTY_NAME))

summary(as.factor(vhist$COUNTY_NAME))

summary(as.factor(year(mdy(vhist_na$ELECTION_DATE))))

summary(as.factor(vhist_na$ELECTION_TYPE))
```
There appear to be *several* different counties here that have NA's in the Voting Method column. Most NAs are in Jefferson. Admittedly, that is the most populous of counties, but it is still about 1 in 40 voters! At first glance, the number of NAs rises exponentially with the totals of voters per county.

Every single NA is concentrated before 2013, when it could be reasonaby supposed that at least local elections were conducted through VBM. The vast, VAST majority are before 2002. 

The distribution of NAs across election types seems fairly consistent with the total amount of votes cast in each contest type. This is reasonable, since it suggests that counting method is independent of election type. 
   

#Data Wrangling for Graphs
   
First create a turnouts table with just raw number of votes per voting method. See file with tables for table on how each voting method was coded.
```{r}
#Moving on with code as if problems were not there

#Sorting out voting metthods
vhist$VOTING_METHOD[vhist$VOTING_METHOD == "Early Voting - DRE" | vhist$VOTING_METHOD == "Early Voting" | 
                      vhist$VOTING_METHOD == "Vote Center" | vhist$VOTING_METHOD == "In Person - DRE" |
                      vhist$VOTING_METHOD == "Polling Place" | vhist$VOTING_METHOD == "Vote Center - DRE"] <- "In Person" 
vhist$VOTING_METHOD[vhist$VOTING_METHOD == "Mail Ballot - DRE"| vhist$VOTING_METHOD == "Absentee Carry" |
                      vhist$VOTING_METHOD == "Absentee Mail"] <- "Mail Ballot" 

#Make date variable
vhist$ELECTION_DATE <- mdy(vhist$ELECTION_DATE)

#Create turnouts file with raw vote counts
turnouts <- vhist %>%
  mutate(count = 1) %>%
  group_by(COUNTY_NAME, ELECTION_DESCRIPTION, ELECTION_TYPE, ELECTION_DATE, VOTING_METHOD) %>%
  summarize(total_votes = sum(count)) %>%
  spread(key = VOTING_METHOD, value = total_votes) %>%
  filter(year(ELECTION_DATE) >= 1992)

#Rename so that no spaces exist
names(turnouts)[5:6] <- c("IN_PERSON", "VBM")

#Replace NA values with 0
turnouts$IN_PERSON[is.na(turnouts$IN_PERSON)] = 0
turnouts$VBM[is.na(turnouts$VBM)] = 0

head(turnouts)
```
    
I then calculate the total amount of registered voters per county. I include active and non-active voters.
```{r, warning=FALSE, error=FALSE}
#Set DATE variables
vrf$REGISTRATION_DATE <- mdy(vrf$REGISTRATION_DATE)
vrf$EFFECTIVE_DATE <- mdy(vrf$EFFECTIVE_DATE)

#Filter inactive or corrupt voters
vrf <- vrf %>%
  filter(year(REGISTRATION_DATE) > 1910 && year(EFFECTIVE_DATE) > 1910)

#Create sums of registrants per year
col_county_registrants <- vrf %>%
  mutate(count = 1, YEAR_REGISTERED = year(REGISTRATION_DATE)) %>%
  group_by(COUNTY, YEAR_REGISTERED) %>%
  summarize(REGISTERED_IN_YEAR = sum(count))

#Create total registrants per year
col_county_registrants <- col_county_registrants %>%
  group_by(COUNTY) %>%
  mutate(REGISTRANTS = cumsum(REGISTERED_IN_YEAR)) %>%
  filter(YEAR_REGISTERED >= 1992) 

head(col_county_registrants)
```
   
#Graph Creation Step
   
I will include four graphs:
*A cumulative graph for the whole of Colorado
*A county level graph including all counties
*A county level graph with only the largest CO counties
*A county level graph, with the largest counties coded a different color than the rest
    
*Statewide Graph*
```{r, fig.align= 'center'}
##Graph of Colorado in general##

#!!!! What are the off year general elections about?? !!!!#

#Total Registered Colorado VOtesr by year
total_registrants <- col_county_registrants %>%
  ungroup() %>%
  group_by(YEAR_REGISTERED) %>%
  summarise(TOTAL_REGISTRANTS = sum(REGISTRANTS))

#Create statewide turnout dataset
turnouts_statewide <- turnouts %>%
  ungroup() %>%
  select(1, 3:6) %>%
  filter(ELECTION_TYPE %in% c("General", "Coordinated", "Primary")) %>%
  mutate(ELECTION_DATE = year(ELECTION_DATE)) %>%
  group_by(ELECTION_TYPE, ELECTION_DATE) %>%
  summarise(IN_PERSON = sum(IN_PERSON), VBM = sum(VBM), TOTAL_VOTES = sum(IN_PERSON, VBM)) 


#Remove even year coordinated
turnouts_statewide <- turnouts_statewide[!(turnouts_statewide$ELECTION_TYPE == "Coordinated" & as.integer(turnouts_statewide$ELECTION_DATE) %% 2 == 0), ]

#Remove odd year Primaries
turnouts_statewide <- turnouts_statewide[!(turnouts_statewide$ELECTION_TYPE == "Primary" & as.integer(turnouts_statewide$ELECTION_DATE) %% 2 != 0), ]

#Remove odd year Generals
turnouts_statewide <- turnouts_statewide[!(turnouts_statewide$ELECTION_TYPE == "General" & as.integer(turnouts_statewide$ELECTION_DATE) %% 2 != 0), ]


 
#Join the data
names(turnouts_statewide)[2] <- "ELECTION_YEAR"
names(total_registrants)[1] <- "ELECTION_YEAR"

turnouts_statewide <- merge(turnouts_statewide, total_registrants, by = "ELECTION_YEAR")

#Make some tweeks 
turnouts_statewide <- turnouts_statewide %>%
  mutate(TURNOUT = TOTAL_VOTES/TOTAL_REGISTRANTS)

turnouts_statewide$ELECTION_TYPE <- as.factor(turnouts_statewide$ELECTION_TYPE)  

#GRAPH!!
ggplot(turnouts_statewide, aes(x = ELECTION_YEAR, y = TURNOUT, col = ELECTION_TYPE)) +
  scale_color_manual(values = c("#2F9395", "Dark Orange", "dark blue")) +
  geom_vline(xintercept = 1992, col = "red") +
  geom_vline(xintercept = 2008, col = "red") +
  geom_vline(xintercept = 2013, col = "red") +
  geom_hline(yintercept = 1) +
  geom_line() +
  geom_point(size = .9) +
  labs(title = "Turnout in Colorado Elections, 1992-2017", x = "Election Year", y = "Turnout as % of Registared Voters", col = "Election Type") +
  theme_minimal()
```
   
   
*County Level Graphs*
```{r, fig.align= 'center'}
#Create County-level turnout file
turnouts_by_county <- turnouts %>%
  ungroup() %>%
  select(1, 3:6) %>%
  filter(ELECTION_TYPE %in% c("General", "Coordinated", "Primary")) %>%
  mutate(ELECTION_DATE = year(ELECTION_DATE))

#Rename in preparation for merger
names(turnouts_by_county)[1:3] <- c("COUNTY","ELECTION_TYPE", "ELECTION_YEAR")
names(col_county_registrants)[2] <- "ELECTION_YEAR"

#Merge
turnouts_by_county <- merge(turnouts_by_county, col_county_registrants, by = c("COUNTY", "ELECTION_YEAR"))

#Tidy data for graphing
turnouts_by_county <- turnouts_by_county %>%
  ungroup() %>%
  mutate(TURNOUT = (VBM + IN_PERSON)/REGISTRANTS) %>%
  select(-6)
  

#Remove even year coordinated
turnouts_by_county <- turnouts_by_county[!(turnouts_by_county$ELECTION_TYPE == "Coordinated" & as.integer(turnouts_by_county$ELECTION_YEAR) %% 2 == 0), ]

#Remove odd year Primaries
turnouts_by_county <- turnouts_by_county[!(turnouts_by_county$ELECTION_TYPE == "Primary" & as.integer(turnouts_by_county$ELECTION_YEAR) %% 2 != 0), ]

#Remove odd year Generals
turnouts_by_county <- turnouts_by_county[!(turnouts_by_county$ELECTION_TYPE == "General" & as.integer(turnouts_by_county$ELECTION_YEAR) %% 2 != 0), ]

#Make County a Factor
turnouts_by_county$COUNTY <- as.factor(turnouts_by_county$COUNTY)

#GRAPH

##Just the first I could pump out
ggplot(turnouts_by_county, aes(x = ELECTION_YEAR, y = TURNOUT, col = COUNTY)) +
  facet_wrap(facets = "ELECTION_TYPE") +
  geom_vline(xintercept = 1992, col = "red") +
  geom_vline(xintercept = 2008, col = "red") +
  geom_vline(xintercept = 2013, col = "red") +
  geom_hline(yintercept = 1) +
  geom_point(size = .2) +
  geom_line(alpha = .3) +
  labs(title = "Turnout in Colorado Elections by County, 1992-2017", x = "Election Year", y = "Turnout as % of Registered Voters") +
  guides(col=FALSE)


##A graph with only the largest 8 counties
big_seven_turnouts <- turnouts_by_county %>%
  filter(COUNTY %in% c("Jefferson", "El Paso", "Denver", "Arapahoe", "Adams", "Larimer", "Boulder", "Douglas"))

ggplot(big_seven_turnouts, aes(x = ELECTION_YEAR, y = TURNOUT, col = COUNTY)) +
  facet_wrap(facets = "ELECTION_TYPE") +
  geom_vline(xintercept = 1992, col = "red") +
  geom_vline(xintercept = 2008, col = "red") +
  geom_vline(xintercept = 2013, col = "red") +
  geom_hline(yintercept = 1) +
  geom_point(size = .2) +
  geom_line(alpha = .8) +
  labs(title = "Turnout in Colorado Elections for the Largest 8 Counties, 1992-2017", x = "Election Year", y = "Turnout as % of Registered Voters")
  
#Note what happens to Jefferson!

 ##Now the same graph, but with the rest of the counties included in grayscale
  grayscale_turnouts <- turnouts_by_county %>%
    mutate(BIG_SEVEN <- ifelse(COUNTY %in% c("Jefferson", "El Paso", "Denver", "Arapahoe", "Adams", "Larimer", "Boulder", "Douglas"), "Big Eight", "Other"))
  
  names(grayscale_turnouts)[8] <- "BIG_EIGHT"
  
  ggplot(grayscale_turnouts, aes(x = ELECTION_YEAR, y = TURNOUT, col = BIG_EIGHT)) +
    scale_color_manual(values = c("dark orange", "grey")) +
    facet_wrap(facets = "ELECTION_TYPE") +
    geom_vline(xi
    geom_vline(xintercept = 200ntercept = 1992, col = "red") +8, col = "red") +
    geom_vline(xintercept = 2013, col = "red") +
    geom_hline(yintercept = 1) +
    geom_point(aes(shape = BIG_EIGHT)) +
    labs(title = "Turnout in Colorado Elections on a County Level, 1992-2017", x = "Election Year", y = "Turnout as % of Registared Voters")
```



