---
title: "CO 2017 Variable Analysis"
author: "Theodore Dounias"
date: "5/10/2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(readr)
library(lubridate)
library(knitr)

CO_2017_VHist_full <- read_csv("data/2017_CO/VHist_2017/CO_2017_VHist_full.csv")
CO_2017_VRF_full <- read_csv("data/2017_CO/VRF_2017/CO_2017_VRF_full.csv")
```

#Voter History

##Dimensions

```{r}
#Voter History Records Dimensions
dim(CO_2017_VHist_full)

#Voter Registration Records Dimensions
dim(CO_2017_VRF_full)
```
    
##Voter History Variables
   
###Voter ID
    
```{r}
#Voter ID NA's
sum(is.na(CO_2017_VHist_full$VOTER_ID))

#Unique Voter ID Values
length(unique(CO_2017_VHist_full$VOTER_ID))
```
   
This last number is smaller than the total amount of entries, which makes sense since multiple records of voting exist for single people.
    
###Election Type
    
```{r}
#Number of NA's
sum(is.na(CO_2017_VHist_full$ELECTION_TYPE))

#Summary of election types
summary(as.factor(CO_2017_VHist_full$ELECTION_TYPE))
```
   
There are 9 types of elections listed: Coordinated, General, Municipal, Municipal Run-Off, Primary, Recall, School, Special, and Special District.
    
###Election Date
   
```{r, fig.align='center'}
#Number of NA's
sum(is.na(CO_2017_VHist_full$ELECTION_DATE))

#Earliest and Latest Date Listed
dates_vhist <- mdy(CO_2017_VHist_full$ELECTION_DATE)

#Earliest Date
min(dates_vhist)

#Latest Date
max(dates_vhist)

#Number of Dates--Not necessarily number of elections!
length(unique(dates_vhist))

dates_vhist <- data.frame(dates_vhist)

#Distribution of Dates
dist_dates <- dates_vhist %>%
  mutate("count" = 1) %>%
  group_by(dates_vhist) %>%
  summarise("Total" = sum(count))

#Plot of all dates with frequencies
ggplot(dist_dates, aes(y = Total)) +
  geom_point(aes(x = dates_vhist))

#Number of invalid dates
kable(dist_dates[1:2,])

#Plot with invalid dates removed
dist_dates <- dist_dates[-(1:2),]

#Plot without invalid dates
ggplot(dist_dates, aes(y = Total)) +
  geom_jitter(aes(x = dates_vhist, col = month(dates_vhist)))
```
   
The plots in this section show how many entries are present for each date. Each point represents a date on the x axis, and the y axis is the number of observations/votes for that date.
   
###Election Description

```{r}
#Different Election Descriptions
summary(as.factor(CO_2017_VHist_full$ELECTION_DESCRIPTION))

#Why is the previous function cutting off?
length(unique(CO_2017_VHist_full$ELECTION_DESCRIPTION))
#Too many values!

#Number of NA's
sum(is.na(CO_2017_VHist_full$ELECTION_DESCRIPTION))
```
    
###Voting Method
   
```{r}
#Number of NA's
sum(is.na(CO_2017_VHist_full$VOTING_METHOD))

#Percentage NA
sum(is.na(CO_2017_VHist_full$VOTING_METHOD))/length(CO_2017_VHist_full$VOTING_METHOD)

#Types of voting methods
summary(as.factor(CO_2017_VHist_full$VOTING_METHOD))
```
   
There is a large number of NA's in this variable, but it is only about 3% of the data; which seems reasonable.
   
###Party

```{r}
#Number of NA's
sum(is.na(CO_2017_VHist_full$PARTY))

#Percentage of NA's
sum(is.na(CO_2017_VHist_full$PARTY))/length(CO_2017_VHist_full$PARTY)

#Counts of Party Affiliations
summary(as.factor(CO_2017_VHist_full$PARTY))
```
   
I assume here that the three letter variables are just concise forms of the whole part names.
   
###County Name

```{r}
#Number of NA's
sum(is.na(CO_2017_VHist_full$COUNTY_NAME))

#Counts of county votes across all elections
summary(as.factor(CO_2017_VHist_full$COUNTY_NAME))
```

###Switching Voters? A fun result if it works (IN PROGRESS)
```{r, eval = FALSE}
#Counting voters that voted in more than one county
switches <- 0
per_voter <- data.frame(unique(CO_2017_VHist_full$VOTER_ID), 0)
names(per_voter) <- c("ID", "Switches")

vote_data <- CO_2017_VHist_full
vote_data <- vote_data %>%
  arrange(VOTER_ID, mdy(ELECTION_DATE)) %>%
  select(2, 8)
##FIX INDEXING!
a <- 1
for(i in 1:15100){
  if(vote_data[i, 1] == vote_data[i+1, 1] && vote_data[i, 2] != vote_data[i+1, 2]) {
    switches <- switches + 1
    per_voter$Switches[a] <- per_voter$Switches[a] + 1 
  }
  if(vote_data[i, 1] != vote_data[i+1, 1]) a <- a+1
}

```
   
##Voter Registration Variable Analysis

###Voter ID
```{r}
#Voter ID NA's
sum(is.na(CO_2017_VRF_full$VOTER_ID))

#Unique Values
length(unique(CO_2017_VRF_full$VOTER_ID))
```
The number of unique voter IDs is the same as the number of observations in the file; this means that every observation has a unique voter ID.

###County
```{r}
#County NA's
sum(is.na(CO_2017_VRF_full$COUNTY))

#Counties and voter distribution
summary(as.factor(CO_2017_VRF_full$COUNTY))
```
   
###Names

```{r}
#NA's in name variables
sum(is.na(CO_2017_VRF_full$FIRST_NAME))

sum(is.na(CO_2017_VRF_full$MIDDLE_NAME))

sum(is.na(CO_2017_VRF_full$LAST_NAME))

sum(is.na(CO_2017_VRF_full$NAME_SUFFIX))
```
  
###Registration Date
```{r}
#Sum of NA values
sum(is.na(CO_2017_VRF_full$REGISTRATION_DATE))

#Sum of values per year
reg_dates <- mdy(CO_2017_VRF_full$REGISTRATION_DATE)

summary(as.factor(year(reg_dates)))
```
   
###Effective Date
```{r}
#Sum of NA values
sum(is.na(CO_2017_VRF_full$EFFECTIVE_DATE))

#Sum of values per year
eff_dates <- mdy(CO_2017_VRF_full$EFFECTIVE_DATE)

summary(as.factor(year(eff_dates)))
```
   
#Last update of file date
```{r}
#Sum of NA values
sum(is.na(CO_2017_VRF_full$LAST_UPDATED_DATE))

#Sum of values per year
upd_dates <- mdy(CO_2017_VRF_full$LAST_UPDATED_DATE)

summary(as.factor(year(upd_dates)))
```


