---
title: "Data Diagnostics"
author: "Theodore Dounias"
date: "9/23/2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
library(tidyverse)
library(lubridate)
library(knitr)
opts_knit$set(root.dir = '/Users/tdounias/Desktop/Reed_Senior_Thesis/Data_and_results/data/CO_12_to_16')
```

##FULL VOTER HISTORY DIAGNOSTIC

```{r read in and format dates}
vhist_full <- read_csv("full_voter_history.csv")

vhist_full$ELECTION_DATE <- mdy(vhist_full$ELECTION_DATE)
```

First I will go through all the variables. I will count unique values, count NAs, see if there is any pattern to the NAs, and clean up when necessary. 
   
```{r voter ID}
#Number of NAs?
sum(is.na(vhist_full$VOTER_ID))

#These can probably be deleted
vhist_full <- vhist_full[!(is.na(vhist_full$VOTER_ID)), ]
#The new vhist is smaller by 8 observations!
```
   
```{r election type}
#Number of NAs?
sum(is.na(vhist_full$ELECTION_TYPE))
#No NAs!

#Checking for "weird" values
summary(as.factor(vhist_full$ELECTION_TYPE))

#All these types are straightforward, nothing to deal with!
```
   
```{r election description}
#Number of NAs?
sum(is.na(vhist_full$ELECTION_DESCRIPTION))
#No NAs!
```
   
```{r election dates}
#Number of NAs?
sum(is.na(vhist_full$ELECTION_DATE))
#No NAs!

#Checking for "weird" values
unique(vhist_full$ELECTION_DATE)

#Nothing too weird, a few elections miscoded in the 1900s
```
   
```{r voting method}
#Number of NAs?
sum(is.na(vhist_full$VOTING_METHOD))
#A BUNCH of NAs! Let's check them out
missing_method <- vhist_full[(is.na(vhist_full$VOTING_METHOD)), ]

#Guess what county?
summary(as.factor(missing_method$COUNTY_NAME))
#Jefferson County!

#Date distribution
ggplot(missing_method, aes(x = ELECTION_DATE)) +
  geom_density()
#Note most if not all of the dates are before 2005. Therefore, 
#if I only study elections after 2012, this should not be an issue. 

#Checking for "weird" values
summary(as.factor(vhist_full$VOTING_METHOD))
#This is very familiar...so should be fine!
```
   
```{r party vhist}
#Number of NAs?
sum(is.na(vhist_full$PARTY))
#A lot of NAs! 
missing_party <- vhist_full[(is.na(vhist_full$PARTY)), ]

#Guess what county?
summary(as.factor(missing_party$COUNTY_NAME))
#Actually not Jefferson...seems evenly distributed

#Date distribution
unique(missing_party$ELECTION_DATE)
#All elections are between 2008-2009! 
#I wonder what happened...

#Checking for "weird" values generally
summary(as.factor(vhist_full$PARTY))
#Apart from double-coding (DEM/DEMOCRAT)
#Everything else seems fine
```
    
Conclusion here is that the only potentially worrying issue is some missing method values post 2012. Else the data look pretty solid.

##VOTER REG FILE DIAGNOSTICS
   
This has the potential to become...repetitive. I will read in all variables, describe what each one is, and then exclusivelly keep the useful ones.
   
```{r read in full data}
reg12 <- read_csv("2012reg2.csv")
reg13 <- read_csv("2013reg2.csv")
reg14 <- read_csv("2014reg2.csv")
reg15 <- read_csv("2015reg2.csv")
reg16 <- read_csv("2016reg2.csv")
```

Currently, I think I conceivably only need the following variables:

*Voter ID
*County
*Registration Date
*Voter Status
*Party
*Gender
*Birth Year
*Precinct Code
    
The section that follow analyze these variables across the five datasets.
   
```{r Voter ID Values}
#Check for NAs
sum(is.na(reg16$VOTER_ID))

sum(is.na(reg15$VOTER_ID))

sum(is.na(reg14$VOTER_ID))

sum(is.na(reg13$VOTER_ID))

sum(is.na(reg12$VOTER_ID))

#There are 5 total, 1 in 2013 and 4 in 2012. 
#All in all nothing to write home about
```

```{r County}
sum(is.na(reg16$COUNTY))

sum(is.na(reg15$COUNTY))

sum(is.na(reg14$COUNTY))

sum(is.na(reg13$COUNTY))

sum(is.na(reg12$COUNTY))

#A bit more concerning, some voters are missing county values! However,
#given the sheer number of total votes that I have at my disposal, 
#8k is not a large number. I think I can disregard.
```

```{r reg date}
sum(is.na(reg16$REGISTRATION_DATE))

sum(is.na(reg15$REGISTRATION_DATE))

sum(is.na(reg14$REGISTRATION_DATE))

sum(is.na(reg13$REGISTRATION_DATE))

sum(is.na(reg12$REGISTRATION_DATE))

#The numbers for 2012-13 are oddly familiar...
#Let's try pulling them out!
missing2013 <- reg13[(is.na(reg13$REGISTRATION_DATE)), ]
missing2012 <- reg12[(is.na(reg12$REGISTRATION_DATE)), ]

#NOTE THE FOLLOWING!
head(missing2012)[, 1:7]
head(missing2013)[, 1:7]

#They have NAs for almost all exept ID! 
#I think this may be due to privacy concerns.

#Let's run some more diagnostics here.
#This command should count how many date entries per year
#contain more characters than necessary
sum(na.omit(!(nchar(reg12$REGISTRATION_DATE) == 10)))

sum(na.omit(!(nchar(reg13$REGISTRATION_DATE) == 10)))

sum(na.omit(!(nchar(reg14$REGISTRATION_DATE) == 10)))

sum(na.omit(!(nchar(reg15$REGISTRATION_DATE) == 10)))

sum(na.omit(!(nchar(reg16$REGISTRATION_DATE) == 10)))
#All are exactly as they should be
```
 
```{r voter status}
sum(is.na(reg16$VOTER_STATUS))

sum(is.na(reg15$VOTER_STATUS))

sum(is.na(reg14$VOTER_STATUS))

sum(is.na(reg13$VOTER_STATUS))

sum(is.na(reg12$VOTER_STATUS))

#Similarly as before, and including those hidden for privacy,
#There is a managable amount of NAs in these statuses. 

#Moving on to counts
#2012
summary(as.factor(reg12$VOTER_STATUS))

#2013
summary(as.factor(reg13$VOTER_STATUS))

#2014
summary(as.factor(reg14$VOTER_STATUS))

#2015
summary(as.factor(reg15$VOTER_STATUS))

#2016
summary(as.factor(reg16$VOTER_STATUS))

#Voter status here clearly includes absentee voters 
#and the country they live in. In fact, filtering
#one of the years we get:
country_status <- reg12[!(reg12$VOTER_STATUS %in% c("Active", "Inactive")), ]

#See the problem!
head(country_status)[,20:25]

#These voters' records have been a bit mishandled.
#The variables seem "shifted" to one side
#However, they again are very few per year, 
#So I think I can ignore them.
```

```{r party}
sum(is.na(reg16$PARTY))

sum(is.na(reg15$PARTY))

sum(is.na(reg14$PARTY))

sum(is.na(reg13$PARTY))

sum(is.na(reg12$PARTY))

#Similarly as before, and including those hidden for privacy,
#There is a managable amount of NAs in these statuses. 

#Moving on to counts
#2012
summary(as.factor(reg12$PARTY))

#2013
summary(as.factor(reg13$PARTY))

#2014
summary(as.factor(reg14$PARTY))

#2015
summary(as.factor(reg15$PARTY))
#THERE IS NO PARTY VARIABLE IN 2015!

#2016
summary(as.factor(reg16$PARTY))

#Same issue as before; some files are mismatched,
#so values have "bled" into eachother.
#Need to be careful when doing any analysis that includes Party

```
   
```{r gender}
sum(is.na(reg16$GENDER))

sum(is.na(reg15$GENDER))

sum(is.na(reg14$GENDER))

sum(is.na(reg13$GENDER))

sum(is.na(reg12$GENDER))

#Similarly as before, and including those hidden for privacy,
#There is a managable amount of NAs in these statuses. 

#I am not going to do counts. The results will be the same.
```
    
The issue that arises is taht nothing seems completelly clean past the first few variables, due to the issue of variable "displacement". I will need to be very careful when using any of them for any purpose, and make sure to tidy first.
