---
title: "Turnout_calculations"
author: "Theodore Dounias"
date: "9/23/2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
library(tidyverse)
library(lubridate)
library(knitr)
opts_knit$set(root.dir = '/Users/tdounias/Desktop/Reed_Senior_Thesis/Data_and_results/data/CO_12_to_16')
```

##TEST METHOD FOR TURNOUTS

```{r make sure data is read in}
county_reg_ref <- read_csv("County_Reg_Per_Year.csv") 

vhist_full <- read_csv("full_voter_history.csv")

vhist_full$ELECTION_DATE <- mdy(vhist_full$ELECTION_DATE)
```

   
Returning now to El Paso, I will outline the method I will use to find turnout. I will then calculate turnout statistics for the 2014 midterms and check against reported figures. Let's check out this method on Monday and find ways to automate it for all counties.
   
I will do the following:

*DENOMINATOR*
*Find all IDs that were registered to vote in 2014. I will do this through the county/id file I just created.
*Match these IDs with those from the 2014 VRF. Filter out any with a registration date prior to 22 days before election day. 
*If calculating turnout over active voters, filter out inactive voters.
*The length of the remaining dataset should be the denominator of turnout.
    
*NUMERATOR*
*Use the full history file and filter for election day and county.
*Match the resulting IDs with the registration file created above (inactive+active). They should all have a match. If any don't, investigate before disregarding.
*The number of rows should be the numerator, and should be equal to ballots cast.

   
I will now apply this step by step to El Paso.
```{r el paso denominator}
#Make base denominator
denom_ep_14 <- county_reg_ref %>%
  select(2, 5) %>%
  filter(COUNTY14 == "El Paso")


#Read in and wrangle registration file
reg14 <- read_csv("2014reg2.csv", 
         col_types = cols_only(VOTER_ID = col_guess(), COUNTY = col_guess(), 
                               VOTER_STATUS = col_guess(), 
                               REGISTRATION_DATE = col_guess()))

reg14$REGISTRATION_DATE <- mdy(reg14$REGISTRATION_DATE)

#Check if all IDs are present in the reg file
sum(!(denom_ep_14$VOTER_ID %in% reg14$VOTER_ID))

#Matching IDs
denom_ep_14 <- left_join(denom_ep_14, reg14, by = "VOTER_ID")

#Filtering invalid dates
denom_ep_14 <- denom_ep_14 %>%
  filter(REGISTRATION_DATE <= as.Date("2014-11-04") - 22)

#I have turnout statistics from the SoS as percentage of
#Registered Voters, so no need for further filtering!

denominator <- nrow(denom_ep_14)
```
    
```{r el paso numerator}
#Make file with votes
num_ep_14 <- vhist_full %>%
  filter(ELECTION_DATE == as.Date("2014-11-04") &
           COUNTY_NAME == "El Paso")

#Is any of them not in the reg file?
sum(!(num_ep_14$VOTER_ID %in% denom_ep_14$VOTER_ID))

#Investigate!
missing_voters <- num_ep_14[!(num_ep_14$VOTER_ID %in% denom_ep_14$VOTER_ID),]

#Similarities in PARTY registration...
summary(as.factor(missing_voters$PARTY))
#None appear to have data on their party registration!

#Do they exist in the 2014 reg file?
sum(!(missing_voters$VOTER_ID %in% reg14$VOTER_ID))
#Most of them do!

#Do they exist in the cumulative county reg file?
sum(!(missing_voters$VOTER_ID %in% county_reg_ref$VOTER_ID))
#All of them do!

#Voting methods are different...
summary(as.factor(missing_voters$VOTING_METHOD))

#Check to see active/inactive status
missing_voters <- left_join(missing_voters, reg14, by = "VOTER_ID")

#Almost all active or NA (or weird stuff I will deal with)
summary(as.factor(missing_voters$VOTER_STATUS))

###HOWEVER
#Proceeding for now without their inclusion...
numerator <- sum(num_ep_14$VOTER_ID %in% denom_ep_14$VOTER_ID)

turnout <- numerator/denominator
```
   
It is therefore obvious that these numbers are quite close by the following table:

```{r table creation}
table_df <- data.frame(c("Calculated", "Reported"), 
                       c(numerator, 231635), c(denominator, 424553), 
                       c(signif(turnout, 3)*100, 54.56))

names(table_df) <- c("", "Ballots Cast", "Total Registered", "Turnout")

kable(table_df)
```

