---
title: "Turnout Calculations"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(knitr)
opts_knit$set(root.dir = '/Users/tdounias/Desktop/Reed_Senior_Thesis/Data_and_results/data/CO_12_to_16')
```
   
##TEST FOR EL PASO
```{r data read in, include=FALSE}
reg16 <- read_csv("2016reg2.csv", 
                  col_types = cols_only(VOTER_ID = col_guess(), COUNTY = col_guess(), 
                                   REGISTRATION_DATE = col_guess(), 
                                   VOTER_STATUS = col_guess()))

reg16$REGISTRATION_DATE <- mdy(reg16$REGISTRATION_DATE)

hist16 <- read_csv("2016hist.csv")

hist16$ELECTION_DATE <- mdy(hist16$ELECTION_DATE)
```

I move on to the actual calculations.
```{r wrangle files for el paso}
reg16_elpaso <- reg16 %>%
  filter(COUNTY == "El Paso")

hist16_elpaso <- hist16 %>%
  filter(COUNTY_NAME == "El Paso")
  
#Number of voters that are not in the registration file in general
sum(!(unique(hist16_elpaso$VOTER_ID) %in% reg16_elpaso$VOTER_ID))

#Remove them from the history file
hist16_elpaso <- hist16_elpaso[(hist16_elpaso$VOTER_ID %in% reg16_elpaso$VOTER_ID),]

#Filter for one specific election
hist16_elpaso_general <- hist16_elpaso %>%
  filter(ELECTION_DATE == as.Date("2016-11-08"))

reg16_elpaso_general <- reg16_elpaso %>%
  filter(REGISTRATION_DATE <= as.Date("2016-11-08") - 22)

t16gen <- signif(nrow(hist16_elpaso_general)/nrow(reg16_elpaso_general), 3)
```
    
Based on Colorado SoS reports, there were 457,329 registered voters for the general of 2016 in El Paso. My results show `r nrow(reg16_elpaso_general)`. These two numbers are approximately equal.  
  
The total number of voters is reported as 327,649. Based on my data, I have `r nrow(hist16_elpaso_general)`. 

My calculation is the following `r t16gen`, which is close to the reported 71.64%.
   
I will now move on to try to create the ID file Paul suggested.

##ID FILE FOR VRF

I will first extract county and voter ID values from the voter registration file. 
```{r counties from 2016}
county_reg_ref <- data.frame(reg16$VOTER_ID, reg16$COUNTY)

#Test for duplicate IDs
nrow(county_reg_ref) - length(unique(county_reg_ref$reg16.VOTER_ID))

#Rename
names(county_reg_ref) <- c("VOTER_ID", "COUNTY16")
```
   
Moving on to 2015 files, while removing 2016 files for memory conservation.

```{r counties from 2015}


reg15 <- read_csv("2015reg2.csv", 
                  col_types = cols_only(VOTER_ID = col_guess(), COUNTY = col_guess()))

#How many IDs exist in the 2015 file that are not in the 2016 file?
sum(!(reg15$VOTER_ID %in% county_reg_ref$VOTER_ID))

#This means that after adding all necessary data from 2016, the length
#of the final df should be:
nrow(county_reg_ref) + 72400

#We add these, and also update counties
county_reg_ref <- left_join(county_reg_ref, reg15, by = "VOTER_ID")

names(county_reg_ref)[3] <- "COUNTY15"

#Now for those that dropped off after 2015
dropped15 <- data.frame(reg15[!(reg15$VOTER_ID %in% county_reg_ref$VOTER_ID),], NA)
dropped15 <- dropped15[,c(1, 3, 2)]
names(dropped15) <- c("VOTER_ID", "COUNTY16", "COUNTY15")

county_reg_ref <- rbind(county_reg_ref, dropped15)

#Note that the number of rows is actually correct!
nrow(county_reg_ref)
```
   
I now repeat this process for all other years.
   
```{r counties from 2014}

reg14 <- read_csv("2014reg2.csv", 
                  col_types = cols_only(VOTER_ID = col_guess(), COUNTY = col_guess()))

#How many IDs exist in the 2015 file that are not in the 2016 OR 2015 file?
sum(!(reg14$VOTER_ID %in% county_reg_ref$VOTER_ID))

#This means that after adding all necessary data from 2014, the length
#of the final df should be:
nrow(county_reg_ref) + 265085

#We add these, and also update counties
county_reg_ref <- left_join(county_reg_ref, reg14, by = "VOTER_ID")

names(county_reg_ref)[4] <- "COUNTY14"

#Now for those that dropped off after 2014
dropped14 <- data.frame(reg14[!(reg14$VOTER_ID %in% county_reg_ref$VOTER_ID),], NA, NA)
dropped14 <- dropped14[,c(1, 3, 4, 2)]
names(dropped14) <- c("VOTER_ID", "COUNTY16", "COUNTY15", "COUNTY14")

county_reg_ref <- rbind(county_reg_ref, dropped14)

#Note that the number of rows is actually correct again!
nrow(county_reg_ref)
```
   
Next year
```{r counties from 2013}

reg13 <- read_csv("2013reg2.csv", 
                  col_types = cols_only(VOTER_ID = col_guess(), COUNTY = col_guess()))

#How many IDs exist in the 2015 file that are not in the 2016 OR 2015 file?
sum(!(reg13$VOTER_ID %in% county_reg_ref$VOTER_ID))

#This means that after adding all necessary data from 2014, the length
#of the final df should be:
nrow(county_reg_ref) + 96779

#We add these, and also update counties
county_reg_ref <- left_join(county_reg_ref, reg13, by = "VOTER_ID")

names(county_reg_ref)[5] <- "COUNTY13"

#Now for those that dropped off after 2013
dropped13 <- data.frame(reg13[!(reg13$VOTER_ID %in% county_reg_ref$VOTER_ID),], NA, NA, NA)
dropped13 <- dropped13[,c(1, 3, 4, 5, 2)]
names(dropped13) <- c("VOTER_ID", "COUNTY16", "COUNTY15", "COUNTY14", "COUNTY13")

county_reg_ref <- rbind(county_reg_ref, dropped13)

#Note that the number of rows is actually correct again!
nrow(county_reg_ref)
```
  
Last year!
```{r counties from 2012}

reg12 <- read_csv("2012reg2.csv", 
                  col_types = cols_only(VOTER_ID = col_guess(), COUNTY = col_guess()))

#How many IDs exist in the 2015 file that are not in the 2016 OR 2015 file?
sum(!(reg12$VOTER_ID %in% county_reg_ref$VOTER_ID))

#This means that after adding all necessary data from 2014, the length
#of the final df should be:
nrow(county_reg_ref) + 176260

#We add these, and also update counties
county_reg_ref <- left_join(county_reg_ref, reg12, by = "VOTER_ID")

names(county_reg_ref)[6] <- "COUNTY12"

#Now for those that dropped off after 2012
dropped12 <- data.frame(reg12[!(reg12$VOTER_ID %in% county_reg_ref$VOTER_ID),], NA, NA, NA, NA)
dropped12 <- dropped12[,c(1, 3, 4, 5, 6, 2)]
names(dropped12) <- c("VOTER_ID", "COUNTY16", "COUNTY15", "COUNTY14", "COUNTY13", "COUNTY12")

county_reg_ref <- rbind(county_reg_ref, dropped12)

#Note that the number of rows is actually correct again!
nrow(county_reg_ref)

#This may be confusing, but note the following:
length(unique(reg12$VOTER_ID)) - length(reg12$VOTER_ID)

length(unique(county_reg_ref$VOTER_ID)) - length(county_reg_ref$VOTER_ID)

#This means that the 2012 registration file contains 11 duplicates of voter IDs!
```
    
Apart from everything I just ran, the following should also be noted:
```{r}
#2012-13 files have some NAs in County
sum(is.na(reg16$COUNTY))

sum(is.na(reg15$COUNTY))

sum(is.na(reg14$COUNTY))

sum(is.na(reg13$COUNTY))

sum(is.na(reg12$COUNTY))

#There is one NA in ID in 2013, 
#four NAs in ID in 2012. These are automatically not included by 
#the left join in my code

sum(is.na(reg16$VOTER_ID))

sum(is.na(reg15$VOTER_ID))

sum(is.na(reg14$VOTER_ID))

sum(is.na(reg13$VOTER_ID))

sum(is.na(reg12$VOTER_ID))
```
    
For future reference, this file will be written as a csv and stored in the Google drive under "interim results"
   
    
##LONG VHIST FILE   
   
For all IDs included in the long voter registration file, I will extract voting histories.
```{r base file creation}
#Make sure I have the reference file for VR
county_reg_ref <- read_csv("County_Reg_Per_Year.csv")

#Load 2016 history file
vhist16 <- read_csv("2016hist.csv")

#Create first image of full voter history file
vhist_full <- vhist16[vhist16$VOTER_ID %in% county_reg_ref$VOTER_ID,]

#Note that all voters from the 2016 history file
#have a corresponding registration. This makes sense, given how
#the registration file was constructed
sum(!(vhist16$VOTER_ID %in% county_reg_ref$VOTER_ID))
```
   
Now I move on to sequentially adding the rest of the histories.
```{r add 2015 history}
#Read in the data
vhist15 <- read_csv("2015hist.csv")

#How many from this file are missing from the registration file?
sum(!(vhist15$VOTER_ID %in% county_reg_ref$VOTER_ID))
#NONE! YAY!

#Now on to adding the extra values to the large dataset
missing15 <- vhist15[!(vhist15$VOTER_ID %in% vhist_full$VOTER_ID), ]

vhist_full <- rbind(vhist_full, missing15)
```
   
```{r add 2014 history}
#Read in the data
vhist14 <- read_csv("2014hist.csv")

#How many from this file are missing from the registration file?
sum(!(vhist14$VOTER_ID %in% county_reg_ref$VOTER_ID))
#NONE! YAY!

#Now on to adding the extra values to the large dataset
missing14 <- vhist14[!(vhist14$VOTER_ID %in% vhist_full$VOTER_ID), ]

vhist_full <- rbind(vhist_full, missing14)
```
  
```{r add 2013 history}
#Read in the data
vhist13 <- read_csv("2013hist.csv")

#How many from this file are missing from the registration file?
sum(!(vhist13$VOTER_ID %in% county_reg_ref$VOTER_ID))
#NONE! YAY!

#Now on to adding the extra values to the large dataset
missing13 <- vhist13[!(vhist13$VOTER_ID %in% vhist_full$VOTER_ID), ]

vhist_full <- rbind(vhist_full, missing13)
```

```{r add 2012 history}
#Read in the data
vhist12 <- read_csv("2012hist.csv")

#How many from this file are missing from the registration file?
sum(!(vhist12$VOTER_ID %in% county_reg_ref$VOTER_ID))
#NONE! YAY!

#Now on to adding the extra values to the large dataset
missing12 <- vhist12[!(vhist12$VOTER_ID %in% vhist_full$VOTER_ID), ]

vhist_full <- rbind(vhist_full, missing12)
```
   
```{r date troubleshoot}
#I noticed that some date observations are weird... 
#So I'm fixing them by cutting to max 10 characters!
vhist_full$ELECTION_DATE <- substr(vhist_full$ELECTION_DATE, 1, 10)
```

...and this is the final voter history file! Similarly, I will write this file as a csv and upload it.

##TEST METHOD FOR TURNOUTS

```{r make sure data is read in}
if(!(exists("county_reg_ref"))) county_reg_ref <- 
  read_csv("County_Reg_Per_Year.csv") 

if(!(exists("vhist_full"))) {
  vhist_full <- read_csv("full_voter_history.csv")
  vhist_full$ELECTION_DATE <- mdy(vhist_full$ELECTION_DATE)
} 
```

   
Returning now to El Paso, I will outline the method I will use to find turnout. I will then calculate turnout statistics for the 2014 midterms and check against reported figures. Let's check out this method on Monday and find ways to automate it for all counties.
   
I will do the following:

*DENOMINATOR*
*Find all IDs that were registered to vote in 2014. I will do this through the county/id file I just created.
*Match these IDs with those from the 2014 VRF. Filter out any with a registration date prior to 22 days before election day. 
*If calculating turnout over active voters, filter out inactive voters.
*The length of the remaining dataset should be the denominator of turnout.
    
*NUMERATOR*
*Use the full history file and filter for election day and county.
*Match the resulting IDs with the registration file created above (inactive+active). They should all have a match. If any don't, investigate before disregarding.
*The number of rows should be the numerator, and should be equal to ballots cast.

   
I will now apply this step by step to El Paso.
```{r el paso denominator}
#Make base denominator
denom_ep_14 <- county_reg_ref %>%
  filter(COUNTY14 == "El Paso") %>%
  select(2, 5)

#Read in and wrangle registration file
reg14 <- read_csv("2014reg2.csv", 
         col_types = cols_only(VOTER_ID = col_guess(), COUNTY = col_guess(), 
                               VOTER_STATUS = col_guess(), 
                               REGISTRATION_DATE = col_guess()))

reg14$REGISTRATION_DATE <- mdy(reg14$REGISTRATION_DATE)

#Check if all IDs are present in the reg file
sum(!(denom_ep_14$VOTER_ID %in% reg14$VOTER_ID))

#Matching IDs
denom_ep_14 <- left_join(denom_ep_14, reg14, by = "VOTER_ID")

#Filtering invalid dates
denom_ep_14 <- denom_ep_14 %>%
  filter(REGISTRATION_DATE <= as.Date("2014-11-04") - 22)

#I have turnout statistics from the SoS as percentage of
#Registered Voters, so no need for further filtering!

denominator <- nrow(denom_ep_14)
```
    
```{r el paso numerator}
#Make file with votes
num_ep_14 <- vhist_full %>%
  filter(ELECTION_DATE == as.Date("2014-11-04") &
           COUNTY_NAME == "El Paso")

#Is any of them not in the reg file?
sum(!(num_ep_14$VOTER_ID %in% denom_ep_14$VOTER_ID))

#Investigate!
missing_voters <- num_ep_14[!(num_ep_14$VOTER_ID %in% denom_ep_14$VOTER_ID),]

#Similarities in PARTY registration...
summary(as.factor(missing_voters$PARTY))
#None appear to have data on their party registration!

#Do they exist in the 2014 reg file?
sum(!(missing_voters$VOTER_ID %in% reg14$VOTER_ID))
#Most of them do!

#Do they exist in the cumulative county reg file?
sum(!(missing_voters$VOTER_ID %in% county_reg_ref$VOTER_ID))
#All of them do!

#Voting methods are different...
summary(as.factor(missing_voters$VOTING_METHOD))

#Check to see active/inactive status
missing_voters <- left_join(missing_voters, reg14, by = "VOTER_ID")

#Almost all active or NA (or weird stuff I will deal with)
summary(as.factor(missing_voters$VOTER_STATUS))

###HOWEVER!###
#Proceeding for now without their inclusion...
numerator <- sum(num_ep_14$VOTER_ID %in% denom_ep_14$VOTER_ID)

turnout <- numerator/denominator
```
   
It is therefore obvious that these numbers are quite close by the following table:

```{r table creation}
table_df <- data.frame(c("Calculated", "Reported"), 
                       c(numerator, 231635), c(denominator, 424553), 
                       c(signif(turnout, 3)*100, 54.56))

names(table_df) <- c("", "Ballots Cast", "Total Registered", "Turnout")

kable(table_df)
```

