---
title: "Estimation of Models—Applied"
author: "Theodore Dounias"
date: "11/5/2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(Matrix)
library(lme4)
library(gamm4)
library(pander)
source("~/Desktop/Reed_Senior_Thesis/riggd/R/utils.R")
setwd("~/Desktop/Reed_Senior_Thesis/Data_and_results/data")

model_sample <- read_csv("model_indiv_sample.csv")

#DATA
load("all_turnouts.RData")
demographics <- read_csv("colorado_demographic_stats_by_county.csv")
names(demographics) <- tolower(names(demographics))

model_dt <- turnouts_county_data(turnout_list)

model_dt$dates <- as.factor(model_dt$dates)
model_dt$types <- as.factor(model_dt$types)

model_dt <- filter(model_dt, !is.na(model_dt$reg))

model_dt <- left_join(model_dt, demographics, by = "county")

model_dt$county <- as.factor(model_dt$county)

cnt_vec <- as.character(unique(model_dt$county))
```

# MCMC Estimation Processes for Multilevel Models

In statistical science a Markov Chain is a sequence of random variables whose value depends on the value of the exact previous random variable. In mathematical terms, this would be a sequence $\theta^{(1)}, \theta^{(2)}, \theta^{(3)}, ..., \theta^{(t)}$ where $\mathbb{P}(\Theta = \theta^{(t)}|\theta^{(n)}) = \mathbb{P}(\Theta = \theta^{(t)})$ for $n \in [1,t-2]$ but $\mathbb{P}(\Theta = \theta^{(t)}|\theta^{(n)})$ is dependent on $\theta^{(n)}$ for $n = t-1$. A Markov Chain Monte Carlo simulation uses Bayesian estimation to update each sequential estimate of $\theta$, leading it to converge to the true value being estimated [@gelman_data_2006].  

Multilevel models can be estimated using MCMC sampling. Indicatively, this appendix presents the construction and coding of two types of MCMC samplers based on the Gibbs algorithm and Metropolis-Hastings algorithm. The code and mathematical derivations are adapted to my models from Gelman and Hill (2006).

## Gibbs Sampler for the County Models
  
The Gibbs algorithm works as follows:

\begin{enumerate}
  \item Choose a number of parallel simulation runs (chains). This number should be relatively low. In this example it is set to 3.
  \item For each chain do the following:
  \begin{enumerate}
    \item Initialize vector of parameters $\Theta^{(0)} = \{\theta^{(0)}_1\, \theta^{(0)}_2\, ..., \theta^{(0)}_n\}$
    \item Choose a number of iterations. For each iteration update every parameter in vector $\Theta^{(n_{iteration})}$, based on the values of vector $\Theta^{(n_{iteration} - 1)}$.
  \end{enumerate}
  \item Evaluate convergence between the chains.
\end{enumerate}

If convergence is poor, repeat for more iterations, or follow diagnostic procedures. These are not specified here, but Gelman and Hill provide a good overview [@gelman_data_2006; @gelman_bayesian_2003]. 

### County Model 1 (Only Random County Effects)

A basic multilevel model with only group-level intercept mixed effects can be written as follows:

$$y_i \ \sim \ N(a_{j[i]}, \sigma^2_y),\ i \in [1, n] \\ a_j \ \sim \ N(\mu_{\alpha}, \sigma^2_{\alpha}), \ j \in [1,J]$$
This specification is slightly different from that presented in Chapter 2. Here $\alpha_{j[i]}$ is the coefficient for the group $j$ that individual $i$ belongs to, $\sigma_y, \sigma_{\alpha}$ the variances of the individual and group level distributions respectively, and $\mu_{\alpha}$ the mean of the group-level distribution. In the case of the most basic county-level model estimated in my thesis (County Model 1), $n = 704$ and $J = 64$. Using Maximum Likelihood Estimation, and given that:

\begin{equation}
  \alpha_j|y, \mu_{\alpha}, \sigma_y, \sigma_{\alpha} \ \sim \ N(\hat{\alpha_j}, V_j)
\end{equation}

we can obtain estimates:

\begin{equation}
\hat{\alpha_j} = \frac{\frac{n_{[j]}}{\sigma^2_y}\bar{y}_{[j]} + \frac{1}{\sigma^2{\alpha}}}{\frac{n_{[j]}}{\sigma^2_y} + \frac{1}{\sigma^2{\alpha}}},\ \ \ \ \  V_j = \frac{1}{\frac{n_{[j]}}{\sigma^2_y} + \frac{1}{\sigma^2{\alpha}}},
\end{equation}

where $n_{[j]}$ is the number of observations for group j, and $\bar{y}_{[j]}$ is the mean response for group j. Using these estimates and the common MLE estimates for variance and mean in a normal distribution, it is possible to construct a Gibbs sampler for model coefficients and errors. Step 2(b) in the Gibbs sampler would then be:  
  
\begin{enumerate}
  \item Estimate $a_j, \ j\in[1,J]$ using equations (1), (2).
  \item Estimate $\mu_{\alpha}$ by drawing from $N(\frac{1}{J}\sum_{1}^{J}\alpha_j, \sigma_{\alpha}^2/J)$ using the previous values estimated in step 1.
  \item Estimate $\sigma_y^2$ as $\frac{\frac{1}{n}\sum_{1}^{n}(y_i - \alpha_{j[i]})^2}{X_{n-1}^2}$ where $X_{n-1}^2$ is a draw from a $\chi^2$ distribution with $n-1$ degrees of freedom.
  \item Estimate $\sigma_{\alpha}^2$ as $\frac{\frac{1}{J}\sum_{1}^{J}(\alpha_j - \mu_{\alpha})^2}{X_{J-1}^2}$ where $X_{n-1}^2$ is a draw from a $\chi^2$ distribution with $J-1$ degrees of freedom.
\end{enumerate}

While each step here seems relatively intuitive, the derivations behind some of the details (like the chi-squared distribution) are complex MLE processes and beyond the scope of this thesis. The R code for this algorithm is as follows:

```{r}
## Gibbs sampler in R
a.update <- function(){
  a.new <- rep (NA, J)
  for (j in 1:J){
    n.j <- sum (model_dt$county==cnt_vec[j])
    y.bar.j <- mean (model_dt$turnout[model_dt$county==cnt_vec[j]])
    a.hat.j <- ((n.j/sigma.y^2)*y.bar.j + (1/sigma.a^2)*mu.a)/
               (n.j/sigma.y^2 + 1/sigma.a^2)
    V.a.j <- 1/(n.j/sigma.y^2 + 1/sigma.a^2)
    a.new[j] <- rnorm (1, a.hat.j, sqrt(V.a.j))
  }
  return (a.new)
}
mu.a.update <- function(){
  mu.a.new <- rnorm (1, mean(a), sigma.a/sqrt(J))
  return (mu.a.new)
}
sigma.y.update <- function(){
  sigma.y.new <- sqrt(sum((model_dt$turnout-a[model_dt$county])^2)/rchisq(1,703))
  return (sigma.y.new)
}
sigma.a.update <- function(){
  sigma.a.new <- sqrt(sum((a-mu.a)^2)/rchisq(1,J-1))
  return (sigma.a.new)
}

J <- 64
n.chains <- 3
n.iter <- 1000
sims <- array (NA, c(n.iter, n.chains, J+3))
dimnames (sims) <- list (NULL, NULL, c (paste ("a[", 1:J, "]", sep=""), "mu.a",
   "sigma.y", "sigma.a"))

for (m in 1:n.chains){
  mu.a <- rnorm (1, mean(model_dt$turnout), sd(model_dt$turnout))
  sigma.y <- runif (1, 0, sd(model_dt$turnout))
  sigma.a <- runif (1, 0, sd(model_dt$turnout))
  for (t in 1:n.iter){
    a <- a.update ()
    mu.a <- mu.a.update ()
    sigma.y <- sigma.y.update ()
    sigma.a <- sigma.a.update ()
    sims[t,m,] <- c (a, mu.a, sigma.y, sigma.a)
  }
}
```

```{r, echo = FALSE, results = "asis", message = FALSE}
#Get Values after convergance
vals <- data.frame(rep(0,3))
for(i in 1:67){
  for(j in 1:3){
    vals[j,i] <- mean(sims[-c(1:500),j,i])
  }
}

vals <- vals %>%
  summarise_all(mean)

names(vals) <- c (paste ("a[", 1:J, "]", sep=""), "mu.a",
   "sigma.y", "sigma.a")


est_table <- data.frame(c("Sampler", "Model"), rbind(vals[1,65:67], c(0.469, 0.199, 0.039)))

names(est_table)[1] <- "Calculated from..."

pandoc.table(est_table, justify = c("left", "center", "center", "center"), caption = "Gibbs sampler results for County Model 1 \\label{tab:gibbs_1}")
```

As is obvious from Table, the Gibbs sampler produces values very similar to the ones given by an \textit{R} call of Model 1.

### County Model 2 (Random County Effects and County-Level Predictors)

With slight changes from the previous model the following is the mathematical expression for a mixed effects model with group-level predictors:

$$y_i \ \sim \ N(a_{j[i]}, \sigma^2_y),\ i \in [1, n] \\ a_j \ \sim \ N(U_j\gamma, \sigma^2_{\alpha}), \ j \in [1,J], $$

where $U_j$ is a vector of predictor values for group $j$, and $\gamma$ a vector of group-level coefficients, with the rest of the parameters having the same designation as previously. Bear in mind that the second of the previous expressions can also be written as: 

\begin{equation}
\alpha_j = U_j\gamma + \eta_j, \ \ \eta_j \ \sim \ N(0, \sigma_{\alpha}^2)
\end{equation}


Updating the estimates used previously, it is again possible to construct a Gibbs sampler for model coefficients and errors. Step 2(b) in the Gibbs sampler in this case is:

\begin{enumerate}
  \item Estimate $a_j, \ j\in[1,J]$. Start by calculating $y_i^{temp} = y_i - U_{j[i]}\gamma$. Then calculate an estimate $\hat\eta_j$ and variance matrix $V_j$ from equations (1), (2), by replacing $\hat\alpha_j$ with $\hat\eta_j$ and $y$ with $y^{temp}$. Use $\eta_j \ \sim \ N(\hat\eta_j, V_j)$ to draw errors $\eta_j$ and then use (3) to estimate $\alpha_j$ for $j \in [1,J]$.
  \item Estimate $\gamma$ by first regressing $\alpha$ by predictor matrix $U$ to obtain $\hat\gamma$ and variance matrix $V_{\gamma}$. Then use distribution $\gamma_j \ \sim \ N(\hat\gamma_j, V_j)$ to obtain estimates for vector $\gamma$.
  \item Estimate $\sigma_y^2$ as $\frac{\frac{1}{n}\sum_{1}^{n}(y_i - \alpha_{j[i]})^2}{X_{n-1}^2}$ where $X_{n-1}^2$ is a draw from a $\chi^2$ distribution with $n-1$ degrees of freedom.
  \item Estimate $\sigma_{\alpha}^2$ as $\frac{\frac{1}{J}\sum_{1}^{J}(\alpha_j - U_j\gamma)^2}{X_{J-1}^2}$ where $X_{n-1}^2$ is a draw from a $\chi^2$ distribution with $J-1$ degrees of freedom.
\end{enumerate}

County Model 2, as presented in Chapter 4, includes two county-level predictors: percentage of white residents and percentage of urban population; this means that $U = \{x^{\%white}, x^{\%urban}\}$. Keeping this in mind the following code estimates the coefficients and standard errors for Model 2:

```{r}
## Gibbs sampler for a multilevel model with county predictors

a.update <- function(){
  y.temp <- y - U[county]%*%g
  eta.new <- rep (NA, J)
  for (j in 1:J){
    n.j <- sum (county==j)
    y.bar.j <- mean (y.temp[county==j])
    eta.hat.j <- ((n.j/sigma.y^2)*y.bar.j/
                 (n.j/sigma.y^2 + 1/sigma.a^2))
    V.eta.j <- 1/(n.j/sigma.y^2 + 1/sigma.a^2)
    eta.new[j] <- rnorm (1, eta.hat.j, sqrt(V.eta.j))
  }
  a.new <- U%*%g + eta.new
  return (a.new)
}

g.update <- function(){
  lm.0 <- lm (a ~ U)
  g.new <- sim (lm.0, n.sims=1)
  return (g.new)
}
sigma.y.update <- function(){
  sigma.y.new <- sqrt(sum((y-a[county]-X%*%b)^2)/rchisq(1,n-1))
  return (sigma.y.new)
}
sigma.a.update <- function(){
  sigma.a.new <- sqrt(sum((a-U%*%g)^2)/rchisq(1,J-1))
  return (sigma.a.new)
}
```
